---
title: "dplyr 1: key verbs"
subtitle: "Daryn Ramsden"
author: "thisisdaryn at gmail dot com"
date: "last updated: `r Sys.Date()`"
institution: ""
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    chakra: libs/remark.js
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---



```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(fig.align="center", fig.width=5, fig.height=5, warning = FALSE, message = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
duo_accent(
  primary_color = "ivory",
  secondary_color = "#310A31",
  header_font_google = google_font("Roboto", "400"),
  text_font_google   = google_font("Lato", "300"),
  code_font_family = "Fira Code",
  code_font_url = "https://cdn.rawgit.com/tonsky/FiraCode/1.204/distr/fira_code.css",
  header_color = "#f54278",
  title_slide_text_color = "#354a66",
  padding = "15px 15px 25px 15px"
)
```


## dplyr: a package for data manipulation

The data you get is almost in the form you want

--

<tt>dplyr</tt> is an R package that encapsulates many common data manipulation tasks

--

Sometimes you want to: 
--

  * keep only some of the rows

--

  * keep only some of the columns

--

  * adds new columns

--

  * sort data
  
--

  * provide summary statistics

--

<tt>dplyr</tt> has a function for each of these

  
---
## 5 Key tasks/verbs/functions

  * <tt>filter</tt>: keep only some of the rows 


  * <tt>select</tt>: keep only some of the columns 

  * <tt>mutate</tt> adds new columns



  * <tt>arrange</tt>: sort data
  


  * <tt>summarise</tt> (and <tt>group_by</tt>): provide summary statistics 
  
---
## First, let's get some data

```{r message = FALSE, warning = FALSE, eval = FALSE}
library(readr)

cars2018 <- read_csv("data/cars2018.csv")
```

---
## First, let's get some data

```{r message = FALSE, warning = FALSE, eval = FALSE}
library(readr)

cars2018 <- read_csv("data/cars2018.csv")
cars2018
```


---
## First, let's get some data

```{r message = FALSE, warning = FALSE}
library(readr)

cars2018 <- read_csv("data/cars2018.csv")
cars2018
```



---
### What is this data?


--
Information from the US Department of Energy about 2018 model vehicles
--

#### What do (some) of the fields represent?

  * Model: Make and model of the vehicle 

--

  * Displacement: Measure of the engine's size
--

  * Cylinders: Number of cylinders

--

  * Gears: Number of Gears
  
--
  
  * Transmission: <i>Automatic</i>, <i>Manual</i>, or <i>CVT</i> 
  
--

  * MPG: Miles per gallon (highway driving)

  


---
## How do we use dplyr?

--

```{r eval = TRUE}
library(dplyr)
```
--


### Or ...

--
you can load the entire <tt>tidyverse</tt> 

```{r eval = FALSE}
library(tidyverse)
```



---
### If you've never installed it before 

--

You will need to install the package first 

```{r eval = FALSE}
install.packages("dplyr")
```

--
or 

```{r eval = FALSE}
install.packages("tidyverse")
```




---
### Key thing to remember

--

All of these functions have these in common:

--

  * the first input is a data frame 
  
--

  * there are a variable number of following inputs specifying the manipulation you want to do


---
## filter

#### a function for specifying which rows to keep


Example 1: How do we get all cars with manual transmissions?
--
```{r message = FALSE, eval = FALSE}
manual <- filter(cars2018, Transmission == "Manual")
```

---
## filter

#### a function for specifying which rows to keep


Example 1: How do we get all cars with manual transmissions?

```{r message = FALSE, eval = FALSE}
manual <- filter(cars2018, Transmission == "Manual")
manual
```

  
---
## filter

#### a function for specifying which rows to keep


Example 1: How do we get all employees that are salaried?

```{r message = FALSE}
manual <- filter(cars2018, Transmission == "Manual")
manual
```


---
## filter

#### 


Example 2: How do we get all cars with 4 or more cylinders?
--
```{r}
cyl_4 <- filter(cars2018, Cylinders >= 4)
head(cyl_4)
```


---
## filter

#### 

Example 3: How do we get all cars with manual transmissions and 4 or more cylinders?
--
```{r}
manual_cyl4 <- filter(cars2018, Transmission == "Manual", 
                     Cylinders >= 4)

```

---
### Assessment

How many automatic transmission cars have less than 6 gears?


--

```{r}
auto_l6 <- filter(cars2018, Transmission == "Automatic",
                      Gears < 6)
dim(auto_l6)
```





---
## select


#### a function for specifying which columns to keep

Example: How do we keep only the <tt>Model</tt>, <tt>MPG</tt>, <tt>Transmission</tt> and <tt>Cylinders</tt> columns?  

```{r eval = FALSE}
cars_narrow <- select(cars2018, Model, MPG, Transmission, Cylinders)
```

---
## select


#### a function for specifying which columns to keep

Example: How do we keep only the <tt>Model</tt>, <tt>MPG</tt>, <tt>Transmission</tt> and <tt>Cylinders</tt> columns?  

```{r eval = FALSE}
cars_narrow <- select(cars2018, Model, MPG, Transmission, Cylinders)

cars_narrow
```


---
## select


#### a function for specifying which columns to keep

Example: How do we keep only the <tt>Model</tt>, <tt>MPG</tt>, <tt>Transmission</tt> and <tt>Cylinders</tt> columns?  

```{r}
cars_narrow <- select(cars2018, Model, MPG, Transmission, Cylinders)

cars_narrow
```


---
## select

#### Example: we can use the <tt>-</tt> to specify columns we don't want to keep:

--

How do we keep all columns except the <tt>ID</tt>, <tt>Intake</tt>, and <tt>Exhaust</tt>?

```{r eval = FALSE}
cars_alt <- select(cars2018, -ID, -Intake, -Exhaust)
```


---
## select

#### Example: we can use the <tt>-</tt> to specify columns we don't want to keep:



How do we keep all columns except the <tt>ID</tt>, <tt>Intake</tt>, and <tt>Exhaust</tt>?

```{r eval = FALSE}
cars_alt <- select(cars2018, -ID, -Intake, -Exhaust)
cars_alt
```
---
## select

#### Example: we can use the <tt>-</tt> to specify columns we don't want to keep:



How do we keep all columns except the <tt>ID</tt>, <tt>Intake</tt>, and <tt>Exhaust</tt>

```{r}
cars_alt <- select(cars2018, -ID, -Intake, -Exhaust)
cars_alt
```



---
## arrange 

### a function to sort your data

Example: How do we sort the data so that the cars are sorted by MPG?

```{r}
cars_sorted <- arrange(cars2018, MPG)
```



---
## arrange

Example: How do we sort the data so that the cars are sorted by MPG in descending order?

```{r}
cars_desc <- arrange(cars2018, desc(MPG))
```


---
## arrange

You can also sort by more than one variable

Example: How do we sort the cars by Transmission type (alphabetically) with each type sorted by MPG (descending)?


```{r}
cars_transmpg <- arrange(cars2018, Transmission, desc(MPG))
```



--

Question: How can you check that this works?

--

  * One way: Try the <tt>View</tt> function with the data frame you created as input.



---
### Assessment

Fill in the code below to do the following tasks:

  * start with the original data set (all employees)
  * restrict to CVT transmission vehicles
  * remove the <i>FuelInjection</i>, <i>Aspiration</i> and <i>LTC</i> columns
  * order the data in descending order of MPG
    * if cars have the same MPG order them by Model alphabetically 
  
  
```{r eval = FALSE}
____ <- filter(____, ____ == ____)
____ <- select(____, -____, -_____, -_______)
____ <- arrange(____, ____, ____)
```




---
### Assessment solution





```{r}
df <- filter(cars2018, Transmission == "CVT")
df2 <- select(df, -FuelInjection, -Aspiration, -LTC)
df3 <- arrange(df2, desc(MPG), Model)

df3
```

(You don't have to use the same data frame names.)


---
### Assessment solution: alternative


  * You could also have overwritten the data frame at each step.
    * your environment is less cluttered
    * but seeing the intermediate data may help with troubleshooting/debugging your code

```{r}
df <- filter(cars2018, Transmission == "CVT")
df <- select(df, -FuelInjection, -Aspiration, -LTC)
df <- arrange(df, desc(MPG), Model)

df
```



---
## mutate

#### a function to add new columns


Example: Adding a column that indicates whether a car has a fuel efficiency 30 MPG or greater 

```{r eval = FALSE}
cars_30 <- mutate(cars2018, above30 = if_else(MPG > 30, TRUE, FALSE))
```

---
## mutate

#### a function to add new columns


Example: Adding a column that indicates whether a car has a fuel efficiency 30 MPG or greater 

```{r eval = FALSE}
cars_30 <- mutate(cars2018, above30 = if_else(MPG > 30, TRUE, FALSE))

cars30 
```



---
## mutate

#### a function to add new columns


Example: Adding a column that indicates whether a car has a fuel efficiency 30 MPG or greater 

```{r}
cars_30 <- mutate(cars2018, above30 = if_else(MPG > 30, TRUE, FALSE))

head(cars_30)
```


---
## summarise

#### a function for summary statistics


Example: Provide the following summary statistics for the City of Chicago:

  * the total number of cars
  * the average MPG

```{r eval = FALSE}
report <- summarise(cars2018, NumCars = n(),
                    AvgMPG = mean(MPG, na.rm = TRUE))
```

---
## summarise

#### a function for summary statistics


Example: Provide the following summary statistics for the City of Chicago:

  * the total number of cars
  * the average MPG

```{r eval = FALSE}
report <- summarise(cars2018, NumCars = n(),
                    AvgMPG = mean(MPG, na.rm = TRUE))

report
```
---
## summarise

#### a function for summary statistics


Example: Provide the following summary statistics for the City of Chicago:

  * the total number of cars
  * the average MPG

```{r}
report <- summarise(cars2018, NumCars = n(),
                    AvgMPG = mean(MPG, na.rm = TRUE))

report
```
---

## summarise

#### How can we find the following for Manual Transmission cars?

  * the total number of cars
  * the median MPG

--

```{r}
manual_report <- summarise(manual, NumCars = n(),
                           MedMPG = median(MPG, na.rm = TRUE))
```

---

## summarise

#### How can we find the following for Manual Transmission cars?

  * the total number of cars
  * the median MPG



```{r eval = FALSE}
manual_report <- summarise(manual, NumCars = n(),
                           MedMPG = median(MPG, na.rm = TRUE))
manual_report
```


---

## summarise

#### How can we find the following for Manual Transmission cars?

  * the total number of cars
  * the median MPG



```{r}
manual_report <- summarise(manual, NumCars = n(),
                           MedMPG = median(MPG, na.rm = TRUE))
manual_report
```




---


## group_by

#### A function to put the data into groups (before you summarise)


What if we wanted summary statistics for Automatic, Manual and CVT?

```{r}
grouped_cars <- group_by(cars2018, Transmission)
grouped_cars
```


---

#### How is the grouped data frame different from the original?
--

```{r}
grouped_cars
```


  * Notice the information about groups above the data

  * Rows have been put into 3 different groups based on their department
  


---
### So how can we use group_by?


  * If we use summarise on grouped data, we get summary statistics for each group
  

---
### So how can we use group_by?


  * If we use summarise on grouped data, we get summary statistics for each group
  

Example: For each transmission type, what are:
  * the number of cars
  * the average MPG?
  

---
### So how can we use group_by?


  * If we use summarise on grouped data, we get summary statistics for each group
  

Example: For each transmission type, what are:
  * the number of cars
  * the average MPG?
  
```{r}
grouped_cars <- group_by(cars2018, Transmission)
```
 
 
---
### So how can we use group_by?


  * If we use summarise on grouped data, we get summary statistics for each group
  

Example: For each transmission type, what are:
  * the number of cars
  * the average MPG?
  
```{r eval = FALSE}
grouped_cars <- group_by(cars2018, Transmission)
report <- summarise(grouped_cars, NumCars = n(), 
                    AvgMPG = mean(MPG, na.rm = TRUE))
``` 



---
### So how can we use group_by?


  * If we use summarise on grouped data, we get summary statistics for each group
  

Example: For each transmission type, what are:
  * the number of cars
  * the average MPG?
  
```{r eval = FALSE}
grouped_cars <- group_by(cars2018, Transmission)
report <- summarise(grouped_cars, NumCars = n(), 
                    AvgMPG = mean(MPG, na.rm = TRUE))

head(report)
```


---
### So how can we use group_by?


  * If we use summarise on grouped data, we get summary statistics for each group
  

Example: For each transmission type, what are:
  * the number of cars
  * the average MPG?
  
```{r}
grouped_cars <- group_by(cars2018, Transmission)
report <- summarise(grouped_cars, NumCars = n(), 
                    AvgMPG = mean(MPG, na.rm = TRUE))

head(report)
```

---
### Assessment

#### Create a data frame that:
  * contains the number of cars of each transmission type with fuel efficiency 30 MPG or greater
  * is ordered by the number of cars meeting the criterion (descending)

Reorder the following lines of code and fill in the blanks where necessary


```{r eval = FALSE}
____ <- arrange(____, desc(____))
____ <- summarise(____, ____ = n())
____ <- filter(____, ____ > ____)
____ <- group_by(____, ____)
```



---
### Creating piplelines

#### Can we pass output from one function automatically to other functions?

--

Yes. Using the <tt>%>%</tt> operator (which is in the <tt>dplyr</tt> package)

--

How:

  * place the <tt>%>%</tt> between the functions in the pipeline
  
  * omit the first input to any function that is receiving input from the <tt>%>%</tt>
  
--

Example: Create a report with the number of cars with each transmission type and the average MPG for each type

--

```{r}
transmission_rpt <- group_by(cars2018, Transmission) %>%
  summarise(NumCars = n(), AvgMPG = mean(MPG))

transmission_rpt
```


---

### Alternative solution

```{r}
transmission_rpt <- cars2018 %>% group_by(Transmission) %>%
  summarise(NumCars = n(), AvgMPG = mean(MPG))

transmission_rpt
```



---

### Maybe time for other data